A Statistical Dynamical Model to Predict Extreme Events and Anomalous Features in Shallow Water Waves with Abrupt Depth Change

### 2018-11-27 paper1_tKdV/tKdV_theory.tex
DQ_fixFigures
NM

\begin{abstract}
Understanding and predicting extreme events and their anomalous statistics
in complex nonlinear systems is a grand challenge in climate, material,
and neuroscience, as well as for engineering design. Recent controlled
laboratory experiments in weakly turbulent shallow water with abrupt
depth change (ADC) exhibit a remarkable transition from nearly Gaussian
statistics in incoming wave trains before the ADC to outgoing waves
trains after the ADC with extreme anomalous statistics with large
positive skewness of the surface height. Here we develop a statistical
dynamical model to explain and quantitatively predict the above anomalous
statistical behavior as experimental control parameters are varied.
The first step is to use incoming and outgoing truncated Korteweg-de
Vries (TKdV) equations matched in time at the ADC. The TKdV equation
is a Hamiltonian system which induces incoming and outgoing statistical
Gibbs invariant measures which are statistically matched at the ADC.
The statistical matching of the known nearly Gaussian incoming Gibbs
state at the ADC completely determines the predicted anomalous outgoing
Gibbs state, which can be calculated by a simple sampling algorithm, verified
by direct numerical simulations, and successfully captures key features
of the experiment. There is even an analytic formula for the anomalous
outgoing skewness. The strategy here should be useful for predicting
extreme anomalous statistical behavior in other dispersive media in
different settings.
\end{abstract}

@@

Recent laboratory experiments in weakly turbulent shallow water reveal a remarkable transition from Gaussian to anomalous behavior as surface waves cross an abrupt depth change (ADC). 
A normally-distributed incoming wave train, upstream of the ADC, transitions to a highly non-Gaussian outgoing wave train, downstream of the ADC, that exhibits large positive skewness of the surface height and more frequent extreme events \cite{bolles2018anomalous}.
Here we develop a statistical dynamical model to explain and quantitatively
predict this anomalous behavior as experimental control
parameters are varied. The first step is to use incoming and outgoing
truncated Korteweg-de Vries (TKdV) equations matched in time at the
ADC. The TKdV equation is a Hamiltonian system which induces incoming
and outgoing statistical Gibbs invariant measures which are statistically
matched at the ADC. The statistical matching of the known nearly Gaussian
incoming Gibbs state at the ADC completely determines the predicted
anomalous outgoing Gibbs state, which can be calculated by a simple
MCMC algorithm, verified by direct numerical simulations, and successfully
captures key features of the experiment. There is even an analytic
formula for the anomalous outgoing skewness. The strategy here should
be useful for predicting extreme anomalous statistical behavior in
other dispersive media in different settings  \cite{solli2007optical,hohmann2010freak}.

\todo[inline]{If you like, we could cite papers on `Optical rogue waves' and/or rogue waves in microwave systems here. (references added)}

@@

Controlled laboratory experiments were carried out in  \cite{bolles2018anomalous} to examine the statistical behavior of surface waves crossing an ADC. In these experiments, nearly unidirectional waves are generated by a paddle wheel and propagate through a long, narrow wave tank. Midway through, the waves encounter a step in the bottom topography, and thus abruptly transition from one depth to another. The paddle wheel is forced with angle
\[
\theta\left(t\right)=\theta_{0}+\Delta\theta\sum_{n=1}^{N}a_{n}\cos\left(\omega_{n}t+\delta_{n}\right),\;E\sim\left(\Delta\theta\right)^{2}\sum_{n}a_{n}^{2}\omega_{n}^{2}.
\]
where the weights $a_n$ are Gaussian in spectral space. The peak frequency $\omega_p$ gives rise to a characteristic wavelength $\lambda_c$ of water waves, which can be estimated using the dispersion relation.
Above, the total energy $E$ in the system is determined by the angle amplitude
$\Delta\theta$. Optical measurements of the free surface reveal a number of surprising statistical features:
\begin{itemize}
\item Distinct statistics are found between the incoming and outgoing wave
disturbances: incoming waves display near-Gaussian statistics,
while outgoing waves skew strongly towards positive displacement.
\item The degree of non-Gaussianity present in the outgoing wave field depends on the total energy of the system: 
larger driving amplitudes $\Delta\theta$ generate stronger skewness in the surface displacement PDFs.
\item \textcolor{blue}{The waves also show different slopes in the power spectra at upstream and downstream locations}.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1.]{./fig1}

>>>>

\begin{abstract}
Understanding and predicting extreme events and their anomalous statistics
in complex nonlinear systems is a grand challenge in climate, material,
and neuroscience, as well as for engineering design. Recent controlled
laboratory experiments in weakly turbulent shallow water with abrupt
depth change (ADC) exhibit a remarkable transition from nearly Gaussian
statistics in incoming wave trains before the ADC to outgoing waves
trains after the ADC with extreme anomalous statistics with large
positive skewness of the surface height. Here we develop a statistical
dynamical model to explain and quantitatively predict the above anomalous
statistical behavior as experimental control parameters are varied.
The first step is to use incoming and outgoing truncated Korteweg-de
Vries (TKdV) equations matched in time at the ADC. The TKdV equation
is a Hamiltonian system which induces incoming and outgoing statistical
Gibbs invariant measures.
The statistical matching of the known nearly Gaussian incoming Gibbs
state at the ADC completely determines the predicted anomalous outgoing
Gibbs state, which can be calculated by a simple sampling algorithm, verified
by direct numerical simulations, and successfully captures key features
of the experiment. There is even an analytic formula for the anomalous
outgoing skewness. The strategy here should be useful for predicting
extreme anomalous statistical behavior in other dispersive media.
\end{abstract}

@@

Recent laboratory experiments in weakly turbulent shallow water reveal a remarkable transition from Gaussian to anomalous behavior as surface waves cross an abrupt depth change (ADC). 
A normally-distributed incoming wave train, upstream of the ADC, transitions to a highly non-Gaussian outgoing wave train, downstream of the ADC, that exhibits large positive skewness of the surface height and more frequent extreme events \cite{bolles2018anomalous}.
Here we develop a statistical dynamical model to explain and quantitatively
predict this anomalous behavior as experimental control
parameters are varied. The first step is to use incoming and outgoing
truncated Korteweg-de Vries (TKdV) equations matched in time at the
ADC. The TKdV equation is a Hamiltonian system which induces incoming
and outgoing Gibbs invariant measures. The statistical matching of the known nearly Gaussian
incoming Gibbs state at the ADC completely determines the predicted
anomalous outgoing Gibbs state, which can be calculated by a simple
MCMC algorithm, verified by direct numerical simulations, and successfully
captures key features of the experiment. There is even an analytic
formula for the anomalous outgoing skewness. The strategy here should
be useful for predicting extreme anomalous statistical behavior in
other dispersive media in different settings  \cite{solli2007optical,hohmann2010freak}.

@@

Controlled laboratory experiments were carried out in  \cite{bolles2018anomalous} to examine the statistical behavior of surface waves crossing an ADC. In these experiments, nearly unidirectional waves are generated by a paddle wheel and propagate through a long, narrow wave tank. Midway through, the waves encounter a step in the bottom topography, and thus abruptly transition to a shallower depth. The paddle wheel is forced with a pseudo-random signal intended to mimic a Gaussian random sea. In particular, the paddle angle is specified as
%The angle of the paddle wheel $\theta(t)$ is prescribed as a pseudo-random signal that mimics a Gaussian random sea 
\[
\theta\left(t\right)=\theta_{0}+\Delta\theta\sum_{n=1}^{N}a_{n}\cos\left(\omega_{n}t+\delta_{n}\right),\;E\sim\left(\Delta\theta\right)^{2}\sum_{n}a_{n}^{2}\omega_{n}^{2}.
\]
where the weights $a_n$ are Gaussian in spectral space. The peak frequency $\omega_p$ gives rise to a characteristic wavelength $\lambda_c$ of water waves, which can be estimated using the dispersion relation.
Above, the total energy $E$ in the system is determined by the angle amplitude
$\Delta\theta$. Optical measurements of the free surface reveal a number of surprising statistical features:
\begin{itemize}
\item Distinct statistics are found between the incoming and outgoing wave
disturbances: incoming waves display near-Gaussian statistics,
while outgoing waves skew strongly towards positive displacement.
\item The degree of non-Gaussianity present in the outgoing wave field depends on the total energy of the system: 
larger driving amplitudes $\Delta\theta$ generate stronger skewness in the surface displacement PDFs.
\item \textcolor{blue}{The waves also show different slopes in the power spectra at upstream and downstream locations}.

@@

\begin{figure}[h]
\centering
\subfloat[skewness from the Gibbs measure]{\includegraphics[scale=1.]{./fig1_1}}
\vspace{-1em}
\subfloat[matching for downstream $\theta^{+}$]{\includegraphics[scale=1.]{./fig1_2}}
\vspace{-1em}
\subfloat[prediction of downstream skewness]{\includegraphics[scale=1.]{./fig1_3}}
